\documentclass[]{article}
\usepackage[numbers]{natbib}
\usepackage{bibentry}
\usepackage[american]{babel}
\nobibliography*

%opening
\title{Growing Assembly Graphs:\\
	   A Brief Literature Review}
\author{Camille Scott}

\begin{document}

\maketitle

\begin{abstract}
Significant prior work has been done on the statistics
of the sequencing process itself, where the sequencing requirements for sampling
a complete genome are explored; the construction of assembly graphs from sampled 
fragments; and the assembly process itself, which is mostly an exercise in the 
interrogation of these graphs. To better understand our problem, I will review several 
papers that span these topics. The first is one of the original papers describing the
sequencing coverage statistics; the second, a 
detailed treatment of the assembly problem through the lens of assembly
graphs' potentials to fully represent the information in a genome; and the third, an 
examination of percolation theory in probabilistic de Bruijn graphs which serves as a 
useful jumping-off point for our project.
\end{abstract}

\section{Original Project Proposal}



A genome can be represented computationally as a sequence of letters representing 
nucleotide bases, a string over the alphabet {A, C, G, T}. Current technology can’t 
sequence an entire genome at once, but can sequence smaller randomly sampled subsets 
(called reads), usually ranging from 100 to 10k bases, with varying degrees of error 
(1-15\%). The computational challenge is the fragment assembly problem: given a multiset 
of reads from a genome, reconstruct the original genome. Complexity arises from regions 
with repeated patterns, sequencing errors, and multiple chromosomes. Methods to solve 
this problems typically use assembly graphs, where reads are connected when there is an 
overlap between them; the original sequence is estimated by walks through this graph. 

Prior work on the problem has assumed the availability of the complete assembly graph; 
that is, the algorithms are offline. While this approach has been sufficient at the scale 
of sequencing projects thus far, the growing volume of sequence data suggests that an 
online approach will eventually need to be adopted. To that end, we propose to develop an 
arrival model of the assembly graph parameterized by depth of coverage, sequencing error 
rate, and an estimator of the underlying genomic complexity. Such a model would aid in 
the development of new online assembly strategies, providing the theoretical groundwork 
for a new class of assemblers.

Since this is an initial step for many biological experiments, the quality of results is 
essential to any posterior analysis, be it a microbiome study or how a new drug interacts 
with antibodies during treatment. Metagenomics assembly has particular opportunity for 
discovery in this area; in these studies, a diverse community of potentially hundreds of 
thousands of microorganisms are sequenced, with varying levels of genome similarity. The 
assembly graph that results is extremely complex, and existing assemblers do not utilize 
strategies capable of coping with the dozens of terabytes of data. A better model can 
lead to improved methods, especially when facing the increasing complexity and amount of 
data being generated.

\section{Sequencing Statistics}

\textbf{Paper}: \bibentry{roach_random_1995}

Roach's work extended a coverage model originally proposed by Lander and Waterman 
\cite{lander_genomic_1988}; the prior model is accurate at low redundancy, but not at the 
higher redundancies present in shotgun studies. This paper came from the days of 
"strategic sequencing," when 
sequencing technology was still in its infancy and optimizing cost versus information was 
extremely important for genomics projects. While we no longer have this concern due to 
the proliferation of extremely cheap shotgun sequencing, Roach's coverage equations are 
still important for understanding how genomic information increases as additional 
fragments are added, and as such can help us understand how assembly graphs grow as more 
fragments are added.

This model starts by assuming a linear discrete genome $G$, from which $n$ fragments
of length $L$ are generated. Fragments cannot overlap the end of $G$, so there are
$G_e = G-L+1$ available start sites for fragments. Redundancy $R$ is $\frac{nL}{G}$. We 
assume that $T$ bases of a pair of fragments are necessary to detect an overlap. They 
then look at the domain space of spacings between start sites $D_k$. They show that the 
density  function for these lengths is: $$f_{D_k}(x) = n(1 - \frac{1}{G_e})^{n-1}$$

From this, and the observation that a gap occurs at a site $S_k$ iff $D_k > L - T$, they
go on to calculate the probability of a gap following a fragment, where $f_G = 
frac{L-T}{G_e}$ is the effective fractional target coverage per fragment (in other words, 
the contribution 
of each fragment), as:
$$P_{gap} \approx \int_{L-T}^{G}{f_{D_k}(x)dx}=(1-f_G)^n$$

From $P_{gap}$, they then go on to derive: the expected number of gaps
$E(N_{gaps})$, the expected number of islands $E(N_{islands})$, the expected fragments
per island, and several models of the distribution of lengths of the islands.

While this may seem somewhat obtuse at first glance, its importance is due it being
directly translatable to a number of assembly graph models by transforming $P_{gap}$
to the probability of edge arrival. For example, existing de Bruijn graph models
set $L$ to be equal to the $k$-mer size $K + 1$ \cite{bresler_optimal_2013}; the analyses 
follow directly. The models provided by Lander, Waterman, and Roach will then directly 
inform any graph model we build.

\section{Assembly Feasibility}

\textbf{Paper}: \bibentry{bresler_optimal_2013}

This work provides a new take on Roach's model. Roach's model does not consider repeats
in the DNA sequence, which add considerable complexity to assembly graphs and the assembly
process. In the absence of repeats, assembly graphs would be collections of disjoint
components which nodes of degree $0$, $1$, or $2$. Repeats introduce cycles, and more
directly, ambiguity to the ordering of traversals which determine the assembled sequence.
Bresler \emph{et al.} study the \textbf{feasibility} of an assembly, whether 
it is possible to construct the original sequence from a set of reads, and the 
\textbf{optimality} of an algorithm in terms of their ability to reconstruct the original
sequence when it is theoretically feasible. Feasibility is a "measure of the intrinsic 
information each read provides about the DNA sequence," and is closely related to the 
length and frequency of repeats in the original genome. This work features a number of 
plots of the performance of specific graph and assembly paradigms versus the lower bound 
on feasibility for increasing $L$. 

The introduction of repetitive sequence is another step toward our growth model. Assembly 
graphs will not be modeled exactly as traditional random graphs, because there is some 
specific underlying sequence which they aim to represent; the properties of that sequence,
such as its repeat content, will need to form the parameters of out model. It's also 
worth noting that the authors identify a two critical points for increasing fragment 
length at which feasibility is achieved with reasonable coverage depths; we may be able 
to find some relationship between these critical points and the geometric phase 
transition observed in \cite{pell_scaling_2012} and described in the next section.

\section{Percolation in de Bruijn Graphs}
\textbf{Paper}: \bibentry{pell_scaling_2012}

This work was primarily concerned with the employing a probabilistic data structure
for the representation of a de Bruijn graph. Previous de Bruijn graph implementations 
stored $k$-mers exactly, requiring massive memory. Here, the authors use a Bloom
filter \cite{broder_network_2004} to only store the hashed $k$-mers; traversal is 
accomplished by exhaustively checking for the presence of neighbors. Because each node 
can have at most eight neighbors (four in either direction), the additional cost is 
minimal. This advancement radically reduced the memory required to store a de Bruijn 
graph, while introducing a new challenge to contend with: Bloom filters have a false 
positive rate, parameterized by the size of their underlying table and the number of 
elements they contain. In a network sense, this means that some nodes will gain neighbors
corresponding to sequences which do not exist in the original set of fragments.

The practical goal of this work was to divide an extremely large set of fragments into 
approachable chunks. In metagenomics, the set of fragments contains genomes from 
potentially millions of species; in a "perfect" data set, where genome share no sequence 
content and there is no error, each of these genomes would be represented by its own 
connected component in the assembly graph. An effective solution to decomposing the 
metagenome assembly problem is then to find all the disconnected components in the 
assembly graph and divide up the fragments based on the component to which they 
correspond. 
To use the probabilistic de Bruijn graph for this task, it was important to determine 
whether false positives in the data structure could cause otherwise disjoint components 
to become erroneously connected. It was observed that for increasing false positive 
rates, the effect on the connectivity was similar to a geometric phase transition: at 
false positives rates of about $0.18$, the average component size begins to grow rapidly. 
This is first shown via simulation, and then studied with percolation theory. They used a 
site percolation model where the probability $p$ of a site being active is equivalent to 
the false positive rate of the bloom filter.

This site percolation model should be relatively close to our model. However, rather than
taking the false positive rate of the Bloom filter as $p$, we will need to leverage
Roach's model for fragment coverage. Intuitively, we expect to see similar behavior with 
respect to component size with increasing sequencing error as Pell \emph{et al.} did with 
respect to increasing error in the Bloom filter. Between this and Bresler \emph{et al.}'s 
critical points for assembly feasibility, we hope to identify other important phase 
transitions in the assembly graph.

\bibliographystyle{plainnat}
\setcitestyle{numbers, open={[}, close={]}}
\bibliography{assembly-graphs}

\end{document}
